### 2016-04-25 text/Chapter04.tex
Edits (Chapter 4)

In the previous chapter, we outlined a number of systems that capture how computations
access the environment in which they are executed. We identified two kinds of systems --
\emph{flat systems} capturing whole-context properties and \emph{structural systems} capturing
per-variable properties. As we show in Section~X, the systems can be further unified using a
single abstraction, but such abstraction is \emph{less powerful} -- \ie~its generality hides
useful properties that we can see when we consider the systems separately. For this reason, we
discuss \emph{flat coeffects} (Chapter~\ref{ch:flat} and Chapter~\ref{ch:semantics}) and
\emph{structural coeffects} (Chapter~\ref{ch:structural}) separately.

In this chapter, we develop a \emph{flat coeffect calculus} that provides a type system for
tracking per-context properties of context-aware programming languages. The \emph{coeffect
calculus} captures the shared properties of such languages. It is parameterized by a \emph{flat
coeffect algebra} and can be instantiated to track implicit parameters, liveness and number of
required past values in dataflow languages.

To capture contextual properties in full generality, the flat coeffect calculus permits multiple
valid typing derivations for a given term. To resolve the ambiguity arising from such generality,
each concrete context-aware language is also equipped with an algorithm for choosing a unique
typing derivation. In the next chapter, we give operational meaning for concrete coeffect
languages based on the flat coeffect calculus and we discuss their safety.

@@

\item The coeffect calculus permits multiple typing derivations due to the ambiguity inherent
  in the conextual lambda abstraction rule. Each concrete context-aware language based on the
  coeffect calculus may resolve this ambiguity differently. We show the procedure for choosing
  unique typing derivation for our three examples (Section~\ref{sec:flat-unique}).

\item We discuss syntactic properties of the calculus, covering type-pre\-ser\-vation for call-by-name
  and call-by-value reduction (Section~\ref{sec:flat-syntax}). We also extend the calculus
  with subtyping and pairs (Section~\ref{sec:flat-exts}). These two sections justify the laws
  of the flat coeffect algebra.
\end{itemize}

@@

In the previous chapter, we looked at three examples of systems that track whole-context
properties. The type systems for whole-context liveness (Section~\ref{sec:applications-flat-live})
and whole-context dataflow (Section~\ref{sec:applications-flat-dataflow}) have a similar
structure in two ways. First, their lambda abstraction duplicates the demands. Given a body with
context demands $\cclrd{r}$, the declaration site context \emph{as well as} the function arrow are
annotated with $\cclrd{r}$. Second, the context demands in the type systems are combined using
two different operators (representing sequential and pointwise operations).

The system for tracking of implicit parameters (Section~\ref{sec:applications-flat-impl})
differs. In lambda abstraction, it splits the context demands between the
declaration site and the call site and they use only a single operator on the indices, typically $\cup$.

Despite the differences, the systems fit the same framework. This becomes apparent when
we consider the categorical structure (Section~\ref{sec:flat-semantics}). However, rather than
starting from the semantics, we first explain how the systems can be unified syntactically
(Section~\ref{sec:flat-calculus-lambda}) and then provide the semantics as a justification.

The development in this chapter can be seen as a counterpart to the well-known development of
\emph{effect systems} \cite{effects-gifford}. The Chapter~\ref{ch:semantics} then links \emph{coeffects}
with \emph{comonads} in the same way in which effect systems can be linked with monads
\cite{monad-notions}. The syntax and type system of the flat coeffect calculus follows
a similar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
in the structure of lambda abstraction as discussed in Section~\ref{sec:applications-structure-lam}
(the relationship with monads is further discussed in Section~\ref{sec:semantics-related}).

@@

\subsection{Reconciling lambda abstraction}
\label{sec:flat-calculus-lambda}

Recall the lambda abstraction rules for the implicit parameters system (annotating the context
with sets of required parameters) and the dataflow system (annotating the context with the
number of past required values):
%
\begin{equation*}
\tyrule{param}
  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \cup \cclrd{s}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
\;
\tyrule{df1}
  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{n}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\cclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{n}} \tau_2 }
\end{equation*}
%
In order to capture both systems using a single calculus, we need a way of unifying the two
systems. For the dataflow system, this can be achieved by over-approximating the number of
required past elements:
%
\begin{equation*}
\tyrule{df2}
  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{\textnormal{min}(n, m)}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\cclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{m}} \tau_2 }
\end{equation*}
%
The rule (\emph{df1}) is admissible in a system that includes the (\emph{df2}) rule.
Furthermore, if we include sub-typing rule (on annotations of functions) and sub-coeffecting rule (on
annotations of contexts), then the reverse is also true -- because
$\cclrd{\textit{min}(n, m)} \leq \cclrd{m}$ and $\cclrd{\textit{min}(n, m)} \leq \cclrd{n}$.
In other words (\emph{df1}) is more precise, but (\emph{df2}) gives a sound over-approximation
with a structure that can be unified with (\emph{param}).

Using a rule such as (\emph{df2}) allows us to give a unified formulation of the flat coeffect
calculus in Section~\ref{sec:flat-calculus-types}, however the coeffect-specific handling of
lambda abstraction remains important in practical implementation in order to obtain a unique
typing derivation for each coeffect program as discussed in Section~\ref{sec:flat-unique}
(and in the implementation in Chatper~\ref{ch:impl}).

@@

The \emph{flat coeffect calculus} is parameterized by a \emph{flat coeffect algebra}, which defines
the structure of context annotations, written as $\cclrd{r}, \cclrd{s}, \cclrd{t}$. These can be
sets of implicit parameters, versions represented as integers or other values.

Typing judgements the attach coeffect annotations $\cclrd{r}, \cclrd{s}, \cclrd{t}$ to the context
and are written as $\coctx{\Gamma}{\cclrd{r}}\vdash e:\tau$.

Before looking at the typing rules, let us clarify how the typing judgements
should be understood. The coeffect calculus provides both analysis of context dependence (type
system) and semantics for context (how it is propagated). These two aspects provide different
ways of reading the judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and the typing rules
used to define it.

\begin{itemize}
\item \textsc{Analysis of context dependence.}
Syntactically, coeffect annotations $\cclrd{r}$ model \emph{context demands}. This means
we can over-approximate them and require more than is actually needed at runtime.

Syntactically, the typing rules are best read top-down. In function application, the context
demands of multiple assumptions (arising from two sub-expressions) are \emph{merged}; in
lambda abstraction, the demands of a single expression (the body) are split between
the declaration site and the call site.

Semantically, the typing rules should be read bottom-up. In application, the capabilities
provided to the term $e_1~e_2$ are \emph{split} between the two sub-expressions; in abstraction,
the capabilities provided by the call site and declaration site are \emph{merged} and passed
to the body.
\end{itemize}

The reason for this asymmetry follows from the fact that the context appears in a \emph{negative
position} in the semantic model (Section~\ref{sec:semantics-theory}). It means that we need to be
careful about using the words \emph{split} and \emph{merge}, because they can be read as meaning
exactly the opposite things. To disambiguate, we always use the term \emph{context demands}
when using the syntactic view, especially in the rest of Chapter~\ref{chf:flat}, and
\emph{context capabilities} or just \emph{available context} when using the semantic view,
especially in Chapter~\ref{ch:semantics}.

@@

To make the flat coeffect system general enough, the algebra consists of three operations.
Two of them, $\cseq$ and $\cpar$, represent the \emph{sequential} and \emph{pointwise} composition,
which are mainly used in function application. The third operator, $\czip$ is used in lambda
abstraction and represents \emph{splitting} of context demands (or, semantically, \emph{merging}
of available context capabilities).

In addition to the three operations, we also require two special values used to annotate
variable access and constant access and a relation that defines the ordering.

\begin{definition}
A \emph{\cclrd{flat coeffect algebra}} $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ is a set
$\C$ together with elements $\cunit, \czero \in \C$, relation $\cleq$ and binary operations
$\cseq, \cpar, \czip$ such that $(\C, \cseq, \cunit)$ is a monoid, $(\C, \cpar, \czero)$ is an
idempotent monoid, $(\C, \cleq)$ is a pre-order and $(\C, \czip)$ is a band (idempotent semigroup).
That is, for all $r,s,t\in \C$:
%
\begin{equation*}
\begin{array}{ccc}
r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t &&
\cunit \;\cseq\; r = r = r \;\cseq\; \cunit
\\
r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
r\; \cpar\; r = r~~~~&
\czero \;\cpar\; r = r = r \;\cpar\; \czero
\\
r \;\czip\; (s \;\czip\; t) = (r \;\czip\; s) \;\czip\; t &&
r\; \czip\; r = r \qquad
\\
\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t &&
t\; \cleq\; t \\[0.5em]
\end{array}
\end{equation*}
%
In addition, the following distributivity axioms hold:
\begin{align*}
\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
\end{align*}
\end{definition}

@@

\item For the $\czip$ operation, we require associativity and idempotence. The idempotence
demand makes it possible to duplicate the coeffects and place the same demand on both
call site and declaration site. Using the example from Section~\ref{sec:flat-calculus-lambda},
this guarantees that the rule (\emph{df1}) is not a special case, but can always be derived
from (\emph{df2}). In some cases, the operator forms a monoid with the unit being the greatest
element of the set $\C$.
\end{itemize}

\noindent
It is worth noting that, in some of the systems, the operators $\cpar$ and $\czip$ are the least
upper bound and the greatest lower bounds of a lattice. For example, in dataflow computations, they
are \emph{max} and \emph{min} respectively. However, this duality does not hold for implicit parameters
(we discuss lattice-based formulation of coeffects in Section~\ref{sec:unified-impl-semilattice}.

Using the syntactic reading, the two operators represent \emph{merging} and \emph{splitting} of context
demands -- in the (\emph{abs}) rule, $\czip$ appears in the assumption and the combined context
demands of the body are split between two positions in the conclusions; in the (\emph{app}) rule,
$\cpar$ appears in the conclusion and combines two context demands from the assumptions.

@@

The flat coeffect algebra requires a pre-order relation $\cleq$. This will be used to introduce
sub-coeffecting and subtyping in Section~\ref{sec:flat-exts-sub}, but we make it a part of the flat
coeffect algebra, as it will be useful for characterization of different kinds of coeffect calculi.
When the idempotent monoid $(\C, \cpar, \czero)$ also has the commutative property (\ie~forms a
semi-lattice), the $\cleq$ relation can be defined as the ordering of the semi-lattice:
%
\begin{equation*}
\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}
\end{equation*}
%
This definition is consistent with all three examples that motivate flat coeffect calculus, but
it cannot be used with the structural coeffects (where it fails for the bounded reuse
calculus) and so we choose not to use it.

@@

\subsection{Language syntax}
\label{sec:flat-calculus-syntax}

The expressions of
the calculus are those of the $\lambda$-calculus with \emph{let} binding. We also include a type
\ident{num} as an example of a concrete base type with numerical constants written as $n$:
%
\begin{equation*}
\begin{array}{rcl}
e &::=& x \sep n \sep \lambda x:\tau.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
\tau &::=& \ident{num} \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
\end{array}
\end{equation*}
%
Note that the lambda abstrction in the syntax is written in the Church-style and requires a type
annotation. This will be used in Section~\ref{sec:flat-unique} where we discuss how to find a
unique typing derivation for context-aware computations. Using Church-style lambda abstraction,
we can directy focus on the more interesting problem of finding unique \emph{coeffect annotations}
rather than solving the problem of type reconstruction.

In the categorically-inspired translation in the next chapter, the function
$\tau_1 \xrightarrow{\cclrd{r}} \tau_2$ is translated into a function
$C^{\cclrd{r}} \tau_1 \rightarrow \tau_2$. However, the type constructor $C^{\cclrd{r}}$
does not itself exist as a syntactical value in the coeffect calculus. This is because we use
comonads to define the \emph{semantics} rather than \emph{embedding} them into the language as in
the meta-language approaches (the distinction has been discussed in Section~\ref{sec:path-sem-langs}).
The annotations $\cclrd{r}$ are formed by an algebraic structure discussed next.

\begin{figure}[t]
\begin{equation*}
\tyrule{var}
  {x : \tau \in \Gamma}
  {\coctx{\Gamma}{\cunit} \vdash x : \tau }
\end{equation*}
\begin{equation*}
\tyrule{const}
  {~}
  {\coctx{\Gamma}{\czero} \vdash n : \ident{num} }
\end{equation*}
\begin{equation*}
\tyrule{app}
  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
   \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 &
    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2}
  {\coctx{\Gamma}{\cclrd{s} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
\end{equation*}

@@

The (\emph{abs}) rule is defined as discussed in Section~\ref{sec:flat-calculus-lambda}. The
body is annotated with context demands $\cclrd{r} \,\czip\, \cclrd{s}$, which are then split
between the context-demands on the declaration site $\cclrd{r}$ and context-demands on
the call site $\cclrd{s}$. Examples of the $\czip$ operator are discussed in the next section.

In function application (\emph{app}), context demands of both expressions and the
function are combined. As discussed in Chapter~\ref{ch:applications}, the pointwise composition
$\cpar$ is used to combine the context demands of the expression representing a function
$\cclrd{r}$ and the context demands of the argument, sequentially composed with the
context-demands of the function $\cclrd{s}\, \cseq \,\cclrd{t}$.

The type system also includes a rule for let-binding. The rule is \emph{not} equivalent to the
typing derivation for $(\lambda x.e_2)~e_1$, but it corresponds to \emph{one} possible typing
derivation. As we show in \ref{sec:flat-exts-let}, the typing used in (\emph{let}) is more
precise than the general rule that can be derived from $(\lambda x.e_2)~e_1$. Additional
constructs such as pairs, sub-coeffecting and sub-typing are covered in Section~\ref{sec:flat-exts}.

@@

The flat coeffect calculus generalizes the flat systems discussed in
Section~\ref{sec:applications-flat} of the previous chapter. We can instantiate it to a specific
use just by providing a flat coeffect algebra. The following summary defines the systems for implicit
parameters, liveness and dataflow. For the latter two, the general calculus has a lambda abstraction
that is compatible with those discussed in Chapter~\ref{ch:flat}, but includes implicit sub-coeffecting.

\begin{example}[Implicit parameters]
Assuming \ident{Id} is a set of implicit parameter names, the flat coeffect algebra
is formed by $(\mathcal{P}(\ident{Id}), \cup, \cup, \cup, \emptyset, \emptyset, \subseteq)$.
\end{example}

>>>>

In the previous chapter, we outlined a number of systems that capture how computations
access the environment in which they are executed. We identified two kinds of systems --
\emph{flat systems} capturing whole-context properties and \emph{structural systems} capturing
per-variable properties. As we show in Section~\ref{sec:unified-unified}, the systems can be
further unified using a single abstraction, but such abstraction is \emph{less powerful} -- \ie~its
generality hides useful properties that we can see when we consider the systems separately. For
this reason, we discuss \emph{flat coeffects} (Chapter~\ref{ch:flat} and Chapter~\ref{ch:semantics})
and \emph{structural coeffects} (Chapter~\ref{ch:structural}) separately.

In this chapter, we develop a \emph{flat coeffect calculus} that provides a type system for
tracking per-context properties of context-aware programming languages. The \emph{coeffect
calculus} captures the shared properties of such languages. It is parameterized by a \emph{flat
coeffect algebra} and can be instantiated to track implicit parameters, liveness and number of
required past values in dataflow languages.
To capture contextual properties in full generality, the flat coeffect calculus permits multiple
valid typing derivations for a given term. To resolve the ambiguity arising from such generality,
each concrete context-aware language is also equipped with an algorithm for choosing a unique
typing derivation. This allows us to explore the language design landscape, while still follow
the usual scoping rules for languages with established approaches (e.g.~implicit parameters in
Haskell).

In the next chapter, we give operational meaning for concrete coeffect
languages based on the flat coeffect calculus and we discuss their safety.

@@

\item The coeffect calculus permits multiple typing derivations due to the ambiguity inherent
  in contextual lambda abstraction. Each concrete context-aware language based on the
  coeffect calculus must specify how such ambiguities are to be resolved. We give the procedure for
  choosing unique typing derivation for our three examples (Section~\ref{sec:flat-unique}).

\item We discuss equational properties of the calculus, covering type-pre\-ser\-vation for call-by-name
  and call-by-value reduction (Section~\ref{sec:flat-syntax}). We also extend the calculus
  with subtyping and pairs (Section~\ref{sec:flat-exts}).
\end{itemize}

@@

In the previous chapter, we looked at three examples of systems that track whole-context
properties. The type systems for whole-context liveness (Section~\ref{sec:applications-flat-live})
and whole-context dataflow (Section~\ref{sec:applications-flat-dataflow}) have a similar
structure in two ways. First, lambda abstraction duplicates their \emph{context demands}. Given a
body with context demands $\cclrd{r}$, the declaration site context \emph{as well as} the function
arrow are annotated with $\cclrd{r}$. Second, the context demands in the type systems are combined
using two different operators (representing sequential and pointwise operations).

The system for tracking implicit parameters (Section~\ref{sec:applications-flat-impl})
differs. In lambda abstraction, it partitions the context demands between the
declaration site and the call site. Furthermore, the operator that combines context demands
is $\cup$ for both sequential and pointwise composition.

Despite the differences, the systems fit the same framework. This becomes apparent when
we consider the categorical structure (Section~\ref{sec:flat-semantics}). Rather than starting
from the categorical semantics, we first explain how the systems can be unified syntactically
(Section~\ref{sec:flat-calculus-lambda}) and then provide the semantics as an additional
justification.

The development in this chapter can be seen as a counterpart to the well-known development of
\emph{effect systems} \cite{effects-gifford}. Chapter~\ref{ch:semantics} then links \emph{coeffects}
with \emph{comonads} in the same way in which effect systems have been linked with monads
\cite{monad-notions}. The syntax and type system of the flat coeffect calculus follows
a similar style as effect systems \cite{effects-polymorphic,effects-talpin-et-al}, but differs
in the structure of lambda abstraction as discussed briefly here and in Section~\ref{sec:applications-structure-lam}
(the relationship with monads is further discussed in Section~\ref{sec:semantics-related}).

@@

\subsection{A unified treatment of lambda abstraction}
\label{sec:flat-calculus-lambda}

Recall the lambda abstraction rules for the implicit parameters coeffect system (annotating
contexts with sets of required parameters) and the dataflow system (annotating contexts with the
number of past required values):
%
\begin{equation*}
\tyrule{param}
  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r} \cup \cclrd{s}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
\;
\tyrule{df1}
  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{n}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\cclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{n}} \tau_2 }
\end{equation*}
%
In order to capture both systems using a single calculus, we need a way of rewriting the
(\emph{df1}) rule such that the annotation in the assumption is in the form $\cclrd{n \disc m}$
for some operation $\cclrd{\disc}$. For the dataflow system, this can be achieved by using
the $\cclrd{\emph{min}}$ function:
%
\begin{equation*}
\tyrule{df2}
  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{\textnormal{min}(n, m)}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\cclrd{n}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{m}} \tau_2 }
\end{equation*}
%
The rule (\emph{df1}) is admissible in a system that includes the (\emph{df2}) rule. That is,
a typing derivation using (\emph{df1}) is also valid when using (\emph{df2}). Furthermore,
if we include sub-typing rule (on annotations of functions) and subcoeffecting rule (on
annotations of contexts), then the reverse is also true -- because
$\cclrd{\textit{min}(n, m)} \leq \cclrd{m}$ and $\cclrd{\textit{min}(n, m)} \leq \cclrd{n}$.
In other words (\emph{df2}) permits an implicit subcoeffecting (and sub-typing) that is not
possible when using the (\emph{df1}) rule, but it has a structure that can be unified
with (\emph{param}).

@@

This section describes the \emph{flat coeffect calculus}. A small programming language based on the
$\lambda$-calculus with a type system that statically tracks context demands. The calculus can
capture different notions of context. The structure of context demands is provided by a
\emph{flat coeffect algebra} (defined in the next section) which is an abstract algebraic structure
that can be instantiated to model concrete context demands (sets of implicit parameters, number
of past values as integers or other information). Annotations that specify context demands are
written as $\cclrd{r}, \cclrd{s}, \cclrd{t}$.

We enrich types and typing judgements with coeffect annotations $\cclrd{r}, \cclrd{s}, \cclrd{t}$;
typing judgements are written as $\coctx{\Gamma}{\cclrd{r}}\vdash e:\tau$. The expressions of
the calculus are those of the $\lambda$-calculus with \emph{let} binding. We also include a type
\ident{num} as an example of a concrete base type with numerical constants written as $n$:
%
\begin{equation*}
\begin{array}{rcl}
e &::=& x \sep n \sep \lambda x:\tau.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2\\
\tau &::=& \ident{num} \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
\end{array}
\end{equation*}
%
Note that the lambda abstrction in the syntax is written in the Church-style and requires a type
annotation. This will be used in Section~\ref{sec:flat-unique} where we discuss how to find a
unique typing derivation for context-aware computations. Using Church-style lambda abstraction,
we can directly focus on the more interesting problem of finding unique \emph{coeffect annotations}
rather than solving the problem of type reconstruction.

In the categorically-inspired translation in the next chapter, the function
$\tau_1 \xrightarrow{\cclrd{r}} \tau_2$ is translated into a function
$C^{\cclrd{r}} \tau_1 \rightarrow \tau_2$. However, the type constructor $C^{\cclrd{r}}$
does not itself exist as a syntactic value in the coeffect calculus. This is because we use
comonads to define the \emph{semantics} rather than \emph{embedding} them into the language as in
the meta-language approaches (the distinction has been discussed in Section~\ref{sec:path-sem-langs}).
The annotations $\cclrd{r}$ are formed by an algebraic structure discussed next.

@@

To make the flat coeffect system general enough, the algebra consists of three operations.
Two of them, $\cseq$ and $\cpar$, represent \emph{sequential} and \emph{pointwise} composition,
which are mainly used in function application. The third operator, $\czip$ is used in lambda
abstraction and represents \emph{splitting} of context demands.

In addition to the three operations, the algebra also requires two special values used to annotate
variable access and constant access and a relation that defines the ordering.

\begin{definition}
A \emph{\cclrd{flat coeffect algebra}} $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ is a set
$\C$ together with elements $\cunit, \czero \in \C$, binary relation $\cleq$ and binary operations
$\cseq, \cpar, \czip$ such that $(\C, \cseq, \cunit)$ is a monoid, $(\C, \cpar, \czero)$ is an
idempotent monoid, $(\C, \czip)$ is a band (idempotent semigroup) and $(\C, \cleq)$ is a pre-order.
That is, for all $r,s,t\in \C$:
%
\begin{equation*}
\begin{array}{ccc}
r \;\cseq\; (s \;\cseq\; t) = (r \;\cseq\; s) \;\cseq\; t &&
\cunit \;\cseq\; r = r = r \;\cseq\; \cunit
\\
r \;\cpar\; (s \;\cpar\; t) = (r \;\cpar\; s) \;\cpar\; t &
r\; \cpar\; r = r~~~~&
\czero \;\cpar\; r = r = r \;\cpar\; \czero
\\
r \;\czip\; (s \;\czip\; t) = (r \;\czip\; s) \;\czip\; t &&
r\; \czip\; r = r \qquad
\\
\textnormal{if}~~r\; \cleq\; s ~~\textnormal{and}~~s\; \cleq\; t~~\textnormal{then}~~r\; \cleq\; t &&
t\; \cleq\; t \\[0.5em]
\end{array}
\end{equation*}
%
In addition, the following distributivity axioms hold:
\begin{align*}
\quad (\cclrd{r}\, \cpar\, \cclrd{s}) \;\cseq\; \cclrd{t} & = (\cclrd{r} \,\cseq\, \cclrd{t}) \;\cpar\; (\cclrd{s}\, \cseq\, \cclrd{t}) \\
\quad \cclrd{t} \;\cseq\; (\cclrd{r}\, \cpar\, \cclrd{s}) & = (\cclrd{t} \,\cseq\, \cclrd{r}) \;\cpar\; (\cclrd{t}\, \cseq\, \cclrd{s})
\end{align*}
\end{definition}

@@

\item For the $\czip$ operation, we require associativity and idempotence. The idempotence
demand makes it possible to duplicate the given coeffects and place the same demand on both
call site and declaration site. Using the example from Section~\ref{sec:flat-calculus-lambda},
this guarantees that the rule (\emph{df1}) is not a special case, but can always be derived
from (\emph{df2}). In some cases, the operator forms a monoid with the unit being the greatest
element of the set $\C$.
\end{itemize}

\noindent
It is worth noting that, in some of the systems, the operators $\cpar$ and $\czip$ are the least
upper bound and the greatest lower bounds of a lattice. For example, in dataflow computations, they
are \emph{max} and \emph{min} respectively. However, this duality does not hold for implicit parameters
(we discuss the lattice-based formulation of coeffects in Section~\ref{sec:unified-impl-semilattice}.

@@

The flat coeffect algebra includes a pre-order relation $\cleq$. This will be used to introduce
subcoeffecting and subtyping in Section~\ref{sec:flat-exts-sub}, but we make it a part of the flat
coeffect algebra, as it will be useful for characterization of different kinds of coeffect calculi.
When the idempotent monoid $(\C, \cpar, \czero)$ is also commutative (\ie~forms a
semi-lattice), the $\cleq$ relation can be defined as the ordering of the semi-lattice:
%
\begin{equation*}
\cclrd{r} \;\cleq\; \cclrd{s} \;\Longleftrightarrow\; \cclrd{r} \;\cpar\; \cclrd{s} \;=\; \cclrd{s}
\end{equation*}
%
This definition is consistent with all three examples that motivate flat coeffect calculus, but
it cannot be used with the structural coeffects (Chapter~\ref{ch:structural}), where it fails for
the bounded reuse calculus. For this reason, we choose not to use it for flat coeffect calculus
either.

@@

\begin{figure}[t]
\begin{equation*}
\tyrule{var}
  {~}
  {\coctx{\Gamma}{\cunit} \vdash x : \tau }\quad(x : \tau \in \Gamma)
\end{equation*}
\begin{equation*}
\tyrule{const}
  {~}
  {\coctx{\Gamma}{\czero} \vdash n : \ident{num} }
\end{equation*}
\begin{equation*}
\tyrule{app}
  {\coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2 &
   \coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_1 }
  {\coctx{\Gamma}{\cclrd{r} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{t})} \vdash e_1~e_2 : \tau_2}
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}\;\czip\;\cclrd{s}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x\!:\!\tau_1.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 }
\end{equation*}
\begin{equation*}
\tyrule{let}
  { \coctx{\Gamma}{\cclrd{r}} \vdash e_1 : \tau_1 &
    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{s}} \vdash e_2 : \tau_2}
  {\coctx{\Gamma}{\cclrd{s} \;\cpar\; (\cclrd{s} \,\cseq\, \cclrd{r})} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2 }
\end{equation*}
%
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau }
  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }\quad(\cclrd{r'} \cleq \cclrd{r})
\end{equation*}
%

@@

The (\emph{abs}) rule is defined as discussed in Section~\ref{sec:flat-calculus-lambda}. The
body is annotated with context demands $\cclrd{r} \,\czip\, \cclrd{s}$, which are then split
between the context-demands on the declaration site $\cclrd{r}$ and context-demands on
the call site $\cclrd{s}$.

In function application (\emph{app}), context demands of both expressions and the
function are combined. As discussed in Chapter~\ref{ch:applications}, sequential composition
is used to combine the context-demands of the argument $\cclrd{s}$ with the context-demands
of the function $\cclrd{t}$. The result $\cclrd{s}\,\cseq\,\cclrd{t}$ is then composed using
pointwise composition with the context demands of the expression that represents the function
$\cclrd{r}$, giving the coeffect $\cclrd{r}\,\cpar\,(\cclrd{s}\cseq\cclrd{t})$.

The type system also includes a rule for let-binding. The rule is \emph{not} equivalent to the
derived rule for $(\lambda x.e_2)~e_1$, but it corresponds to \emph{one} possible typing
derivation. As we show in \ref{sec:flat-exts-let}, the typing used in (\emph{let}) is more
precise than the general rule that can be derived from $(\lambda x.e_2)~e_1$.

To guide understanding of the system, we also show non-syntax-directed (\emph{sub}) rule for
subcoeffecting. The rule states that an expression with context demands $\cclrd{r'}$ can be
treated as an expression with greater context demands $\cclrd{r}$. We return to subcoeffecting,
subtyping and additional constructs such as pairs in Section~\ref{sec:flat-exts}. When
discussing procedure for choocing unique typing in Section~\ref{sec:flat-unique}, we consider
only the syntax-directed part of the system.

Before proceeding, let us clarify how the typing judgements should be understood. The coeffect
calculus can be understood in two ways discussed in this and the next chapter. As a type system
(Chapter~\ref{ch:flat}), it provides analysis of context dependence. As a semantics
(Chapter~\ref{ch:semantics}), it specifies how context is propagated. These two readings provide
different ways of interpreting the judgements $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ and the
typing rules used to define it.

\begin{itemize}
\item \textsc{Analysis of context dependence.}
Syntactically, coeffect annotations $\cclrd{r}$ model \emph{context demands}. This means
we can over-approximate them and require more in the type system than is needed at runtime.

Syntactically, the typing rules are best read top-down (from assumptions to the consequent).
In function application, the context demands of multiple assumptions (arising from two
sub-expressions) are \emph{merged}; in lambda abstraction, the demands of a single expression
(the body) are split between the declaration site and the call site.

Semantically, the typing rules should be read bottom-up (from the consequent to assumptions).
In application, the capabilities provided to the term $e_1~e_2$ are \emph{split} between the
two sub-expressions; in abstraction, the capabilities provided by the call site and declaration
site are \emph{merged} and passed to the body.
\end{itemize}

\noindent
For example, using the syntactic reading, the operators $\czip$ and $\cpar$ represent \emph{merging}
and \emph{splitting} of context demands -- in the (\emph{abs}) rule, $\czip$ appears in the assumption
and the combined context demands of the body are split between two positions in the conclusions;
in the (\emph{app}) rule, $\cpar$ appears in the conclusion and combines two context demands
from the assumptions.

The reason for this asymmetry follows from the fact that the context appears in a \emph{negative
position} in the semantic model (Section~\ref{sec:semantics-theory}). It means that we need to be
careful about using the words \emph{split} and \emph{merge}, because they can be read as meaning
exactly the opposite things. To disambiguate, we always use the term \emph{context demands}
when using the syntactic view, especially in the rest of Chapter~\ref{ch:flat}, and
\emph{context capabilities} or just \emph{available context} when using the semantic view,
especially in Chapter~\ref{ch:semantics}.

@@

The flat coeffect calculus generalizes the three flat systems discussed in
Section~\ref{sec:applications-flat} of the previous chapter. We can instantiate it to a specific
use just by providing a flat coeffect algebra.

\begin{example}[Implicit parameters]
\label{ex:flat-impl}
Assuming \ident{Id} is a set of implicit parameter names written $\ident{?p}$, the flat coeffect
algebra is formed by $(\mathcal{P}(\ident{Id}), \cup, \cup, \cup, \emptyset, \emptyset, \subseteq)$.
\end{example}
